---
title: "CIS 351 Final Project"
author: "Carter Altman, Samantha Guynn, Shane Hall"
format:
    html:
        embed-resources: true
---

```{r}
#| messsage: false

library(tidyverse)
library(janitor)
library(tinytable)
library(ggpubr)
```

# 1. Introduction

We use water quality data collected from five sites in the St. Augustine, Florida area: Camachee, Conch, Shipyard, Oasis, and City. The dataset contains measurements of several environmental variables including Turbidity (NTU), Fecal Coliform (CFU), Chlorophyll (nu/l), Refined Oil (RFU), Temperature (C), Dissolved Oxygen (mg/l), and pH. The motivation for this analysis is to assess and compare water quality conditions across the five sampling sites.

For this project, we focus on the following variables:

- **Numerical:** Turbidity (NTU) and Refined Oil (RFU)
- **Categorical:** Site and Turbidity Category (Low, Moderate, High)

# 2. Import Data

```{r}
water_quality <- read_csv("https://raw.githubusercontent.com/caltman05/water_quality_data_analysis/refs/heads/main/data_9.csv")

water_quality
```

# 3. Data Cleaning and Preprocessing

```{r}
water_quality <- water_quality |>
  clean_names()

water_quality
```

```{r}
water_quality <- water_quality |>
  filter(turbidity_ntu <= 50)

water_quality
```

```{r}
water_quality <- water_quality |>
  mutate(
    turbidity_category = case_when(
      turbidity_ntu < 5  ~ "Low",
      turbidity_ntu <= 20 ~ "Moderate",
      turbidity_ntu > 20 ~ "High"
    )
  )

water_quality
```

```{r}
summary(water_quality)
```

# 4. Descriptive Statistics

## Numerical Variables

```{r}
water_quality_statistics <- water_quality |>
  summarize(
    mean_turbidity = mean(turbidity_ntu, na.rm = TRUE),
    median_turbidity = median(turbidity_ntu, na.rm = TRUE),
    sd_turbidity = sd(turbidity_ntu, na.rm = TRUE),
    mean_fecal = mean(fecal_coliform_cfu, na.rm = TRUE),
    median_fecal = median(fecal_coliform_cfu, na.rm = TRUE),
    sd_fecal = sd(fecal_coliform_cfu, na.rm = TRUE),
    mean_chlorophyll = mean(chlorophyll_nu_l, na.rm = TRUE),
    median_chlorophyll = median(chlorophyll_nu_l, na.rm = TRUE),
    sd_chlorophyll = sd(chlorophyll_nu_l, na.rm = TRUE),
    mean_oil = mean(refined_oil_rfu, na.rm = TRUE),
    median_oil = median(refined_oil_rfu, na.rm = TRUE),
    sd_oil = sd(refined_oil_rfu, na.rm = TRUE),
    mean_temp = mean(temp_c, na.rm = TRUE),
    median_temp = median(temp_c, na.rm = TRUE),
    sd_temp = sd(temp_c, na.rm = TRUE),
    n = n(),
    .by = site
  ) |>
  tt(theme = "striped")

water_quality_statistics
```

## Categorical Variables

```{r}
table(water_quality$site)
table(water_quality$turbidity_category)
```

# 5. Data Visualization

## Histogram of Turbidity

```{r}
plot_turbidity_hist <- water_quality |>
  ggplot(aes(x = turbidity_ntu)) +
  geom_histogram(
    binwidth = 2,
    color = "blue",
    fill = "skyblue"
  ) +
  labs(
    title = "Distribution of Turbidity",
    x = "Turbidity (NTU)",
    y = "Count"
  )

plot_turbidity_hist
```

## Boxplot of Turbidity by Site

```{r}
plot_turbidity_site <- water_quality |>
  ggplot(aes(x = site, y = turbidity_ntu, fill = site)) +
  geom_boxplot() +
  theme_bw() +
  labs(
    title = "Turbidity by Site",
    x = "Site",
    y = "Turbidity (NTU)"
  )

plot_turbidity_site
```

## Scatter Plot of Turbidity vs Refined Oil

```{r}
plot_turbidity_oil <- water_quality |>
  ggplot(aes(x = turbidity_ntu, y = refined_oil_rfu)) +
  geom_point() +
  geom_smooth(
    method = "lm",
    color = "blue",
    se = FALSE
  ) +
  stat_regline_equation(
    label.x = 2,
    label.y = 20
  ) +
  stat_cor(
    label.x = 2,
    label.y = 23
  ) +
  labs(
    title = "Turbidity vs Refined Oil",
    x = "Turbidity (NTU)",
    y = "Refined Oil (RFU)"
  )

plot_turbidity_oil
```

## Stacked Bar Chart of Turbidity Category by Site

```{r}
plot_turbidity_category_site <- water_quality |>
  filter(!is.na(turbidity_category), !is.na(site)) |>
  ggplot(aes(x = site, fill = turbidity_category)) +
  geom_bar() +
  labs(
    title = "Turbidity Category by Site",
    x = "Site",
    y = "Count",
    fill = "Turbidity Category"
  )

plot_turbidity_category_site
```

# 6. Statistical Tests

## ANOVA on `turbidity_ntu` by `site`

From the boxplot in Section 5, we observe that the dataset includes more than two sites. Therefore, we proceed with an ANOVA to determine whether there are significant differences in mean turbidity across sites.

Let $\mu_1, \mu_2, \mu_3, \mu_4, \mu_5$ denote the mean turbidity of Camachee, City, Conch, Oasis, and Shipyard, respectively. The hypotheses are:

- $H_0$: $\mu_1 = \mu_2 = \mu_3 = \mu_4 = \mu_5$
- $H_A$: At least one site has a mean turbidity that differs from the others

```{r}
test_anova <- aov(
  turbidity_ntu ~ site,
  data = water_quality
)

summary(test_anova)
```

Since the $p$-value is $p = 6.84 \times 10^{-9} < 0.05$, we reject $H_0$ and conclude that at least one site has a mean turbidity that differs from the others.

## Kruskal-Wallis Test on `refined_oil_rfu` by `site`

We use the Kruskal-Wallis test as a non-parametric alternative to compare the distribution of refined oil readings across the five sites.

- $H_0$: The distribution of refined oil is the same across all sites
- $H_A$: At least one site has a refined oil distribution that differs from the others

```{r}
test_kruskal <- kruskal.test(
  refined_oil_rfu ~ site,
  data = water_quality
)

test_kruskal
```

Since the $p$-value is $p = 0.292 > 0.05$, we fail to reject $H_0$ and conclude that there is no significant evidence that the distribution of refined oil differs across sites.

## $\chi^2$-Test of Independence on `turbidity_category` and `site`

- $H_0$: `turbidity_category` and `site` are independent
- $H_A$: `turbidity_category` and `site` are dependent

```{r}
test_chisq <- chisq.test(
  table(
    water_quality$turbidity_category,
    water_quality$site
    )
)

test_chisq
```

Since the $p$-value is $p = 1.366 \times 10^{-5} < 0.05$, we reject $H_0$ and conclude that `turbidity_category` and `site` are dependent.

# 7. Conclusion

Based on traffic and visibility at each marina, our data finidngs are consistent with our hypothesis that high boat traffic within the marina affects overall water quality, specificially turbidity.
